---
title: "Supplement to: *Intermediate abundance promotes speciation when dispersal is limited*"
format:
  pdf:
    number-sections: true
    keep-tex: true
    include-in-header: 
      text: |
        \renewcommand{\thefigure}{S\arabic{figure}}
        \renewcommand{\thesection}{S\arabic{section}}
        \renewcommand{\sfdefault}{\rmdefault}
crossref:
  fig-title: "Supplementary Figure"
  fig-prefix: "Supplementary Figure"
csl: oikos.csl
bibliography: abundolism.bib
---


## Computation set-up {#sec-setup}

We make use of a custom R package *abundolism* which, for review purposes, is hosted on anonymous github: [https://anonymous.4open.science/api/repo/abundolism-2DC5](https://anonymous.4open.science/api/repo/abundolism-2DC5)

It can be installed from anonymous github, but note that dependencies on *Rcpp* and *RcppArmadillo* can lead to installation challenges on some infrastructure configurations.

```{r}
#| label: ab-setup
#| eval: false

tmp <- tempfile(fileext = ".zip")
download.file(
    "https://anonymous.4open.science/api/repo/abundolism-2DC5/zip",
    destfile = tmp
)

remotes::install_local(tmp)
```

The following R computing environment is used for all simulations and analyses:

```{r}
#| label: setup

library(abundolism)
library(dplyr)
library(ggplot2)
library(ggpointdensity)
library(cowplot)
library(knitr)

sessionInfo() |> 
    print(locale = FALSE)

```

The quarto [@quarto] document used to generate this supplement can be found with the R command `system.file("ab_supp.qmd", package = "abundolism")`.


## Additional simulation details {#sec-sim}

The birth-death-immigration-speciation simulation tracks the following:

- all local population sizes
- running mean population size per iteration 
- time relative to all rates (i.e. not simply number of iterations)
- whether an incipient speciation event has occurred
- the wait time to full speciation
- whether full speciation has occurred

In each iteration of the simulation an event is chosen at random. Possible events for each local population are

- local birth
- local death
- local speciation
- local immigration

There is also one possible global event: immigration from the source pool.

If there are $n_p$ local populations then the total number of possible events are $n_p \times 4 + 1$. Which event occurs during any given simulation iteration is chosen at random according to the probability of each event. With a local population size of $x_i$, the probability of each event is proportional to its respective rate:


\begin{align}
\mathbb{P}(\text{birth in pop. } i) &= \frac{\lambda x_i}{R} \\
\mathbb{P}(\text{death in pop. } i) &= \frac{\mu x_i}{R} \\
\mathbb{P}(\text{speciation in pop. } i) &= \frac{\nu x_i}{R} \\
\mathbb{P}(\text{local immigration from pop. } i) &= \frac{m_{prop}\gamma x_i}{R} \\
\mathbb{P}(\text{global immigration}) &= \frac{\gamma}{R}
\end{align}


where $R = \gamma + \left( \sum_i^{n_p} x_i \right) (\lambda + \mu + \nu + m_{prop} \gamma)$ is a normalization constant.

In the case of immigration (both local and from the source pool) the recipient of the dispersing individual must also be chosen. We choose the recipient local population at random with equal probability across local populations.

Once an event is randomly sampled, the variables being tracked are updated. Birth and immigration (local or global) augment a local population by 1. Death decrements a local population by 1. Immigration events have no impact on the source of the dispersing individual. Time is updated following the Gillespie algorithm: the sojourn time between events is a random variable with an exponential probability distribution with rate R: $s \sim Exp(R)$. Each time step a random sojourn time is sampled and added to the running total time since the start of the simulation.

If a population experiences local extirpation all of its speciation-related information is re-set including removing any incipient speciation and restoring the wait time to full speciation back to the default of $\tau$.

If full speciation is achieved the simulation stops. If no speciation occurs but the specified number of simulation iterations is reached, the simulation stops. Once stopped, the simulation returns the tracked variables of time, mean population size, and whether full speciation has occurred.



## Simulation experiment details


Simulation parameters are drawn from uniform distributions with the following minimum and maximum limits:

```{r}
#| label: par-range
#| cache: true

par_range <- data.frame(
    param = c("la", "mu", "g", "m_prop", "nu", "tau", "xi"), 
    min   = c(0.1,  0.1,  0.1,  0,        0,     20,  10), 
    max   = c(4,    4,    1,    0.1,      0.1,  200, 100)
)


# fixed params
np <- 10
nstep <- 10000

# number of simulation replicates
nrep <- 5000

# format param names for printing
mutate(par_range, 
       param = sprintf("`%s`", param)) |> 
    kable(format.args = list(scientific = FALSE))
```

The parameters governing number of populations (`np`) and number of time steps (`nstep`) do not vary across simulations and are set to `np =` `r np` and `nstep =` `r nstep`. We simulate a total of `nrep =` `r nrep` replicates, each with a unique, randomly sampled set of parameter values.


### Running the simulation experiment

Parameter values for each simulation replicate are stored in a `data.frame` with one column for each parameter.

```{r}
#| label: par-sample
#| cache: true

# change or remove for different results
set.seed(123)

# data.frame of randomly sampled param values
# one param per column
pars <- sapply(par_range$param, function(p) {
    runif(nrep, 
          par_range[par_range$param == p, "min"], 
          par_range[par_range$param == p, "max"])
}) |> 
    as.data.frame()

```

With those parameters sampled, we can now run the simulation, one realization for each parameter combination. For details on inputs and outputs of the simulation function `sim_BDI_spec` see the help documentation accessible with `?sim_BDI_spec`.

```{r}
#| label: run-sim
#| cache: true
#| dependson: [par-range, par-sample]

# now run the simulation
sim_dat <- sim_BDI_spec(la = pars$la, mu = pars$mu, g = pars$g, 
                        m_prop = pars$m_prop, nu = pars$nu, 
                        tau = pars$tau, xi = pars$xi, 
                        np = np, nstep = nstep, w = 100)

```

Simulation outputs are analyzed in binned windows of timesteps, in this case `w = 100` steps in width. Bins are non-overlapping.


## Interpreting the simulation experiment {#sec-sim-res}

Now we can visually inspect the results. Our main purpose is to study how abundance relates to speciation in this model, but first we check other relationships to ensure the model behaves as expected. In particular, population size should be larger when $\lambda > \mu$ and smaller, supported primarily by immigration. When $\lambda < \mu$ populations will grow indefinitely. When birth and death rates are equal ($\lambda = \mu$) the population could crash or expand depending on small random fluctuations and random immigration. @fig-la-mu shows exactly this pattern. Smaller population sizes for $\lambda > \mu$ likely correspond to time bin windows earlier in the simulation as well as those that stochastically drifted to smaller sizes.

```{r}
#| label: fig-la-mu
#| fig-width: 4
#| fig-height: 3
#| fig-cap: |
#|     Relationship across simulation runs between growth rate ($\lambda - \mu$) 
#|     and mean population size. Panels show simulation runs without and with 
#|     speciation. Points are color coded according to local density of points 
#|     in an attempt to more clearly show the distribution of simulation results.

ggplot(sim_dat, aes(la - mu, mean_pop_size)) +
    geom_hex(data = select(sim_dat, !speciation), 
             mapping = aes(la - mu, mean_pop_size), 
             fill = "gray50") +
    geom_hex() + 
    facet_wrap(vars(speciation), 
               labeller = as_labeller(c(`0` = "No speciation", 
                                        `1` = "Yes speciation"))) +
    scale_y_log10() + 
    xlab(expression(italic(lambda - mu))) + 
    ylab("Mean population size") +
    scale_fill_viridis_c(trans = "log10") +
    theme_cowplot()

```


We also expect that probability of specification should increase with the parameter $\nu$, the rate of incipient speciation. @fig-nu confirms this expectation but shows the relationship to be weak, indicating that while this parameter obviously matters, its effect is modulated by other factors such as abundance and dispersal. This finding is confirmed with logistic regression.

```{r}
#| label: nu-mod
#| message: false

# model of speciation changing with nu
nu_mod <- glm(speciation ~ nu, data = sim_dat, family = binomial)
nu_slope_ci <- confint(nu_mod, 2)


```

The 95% confidence interval on the slope of $\nu$ is (`r round(nu_slope_ci[1], 3)`, `r round(nu_slope_ci[2], 3)`).

With results from the simulation we report confidence intervals rather than hypothesis tests because sample sizes from simulations are arbitrary and thus degrees of freedom can be manipulated to produce statistically significant results from biologically meaningless patterns. By contrast, confidence intervals more meaningfully indicate the magnitude of effects in the simulation experiment.


```{r}
#| label: fig-nu
#| fig-width: 4
#| fig-height: 3
#| fig-cap: |
#|     Relationship across simulation runs between incipient speciation rate
#|     and speciation. Points are jittered around 1 and 0 and are color coded 
#|     according to local density of points all in an attempt to more clearly 
#|     show the distribution of simulation results. Curve is a 
#|     binomial generalize linear model and gray region shows 95% confidence 
#|     envelope. 


logistic_hist_plot(sim_dat, nu, 
                   xlab = expression("Incipient speciation rate"~italic(nu)), 
                   xlog = FALSE,
                   ymin = 0, ymax = 0.008,
                   formula = speciation ~ nu)


```


We also expect dispersal to decrease the chance of speciation because dispersal prolongs the wait time until full speciation. But we have to evaluate this expectation with respect to *relative* dispersal, that is dispersal relative to birth because these rate trade-off in determining whether a population grows due to intrinsic (birth) or extrinsic (dispersal) forces. Dispersal comes from two sources: global dispersal and local dispersal, so we add these two sources when considering total dispersal. @fig-disp Confirms the negative relationship between speciation and dispersal. We again preform a binomial regression of this relationship.

```{r}
#| label: disp-mod
#| message: false

# model of speciation changing with dispersal
disp_mod <- glm(speciation ~ I(log((g + m_prop * g) / la)), 
                data = sim_dat, family = binomial)
disp_slope_ci <- confint(disp_mod)
```

The slope from the logistic regression model for log dispersal normalized by birth rate has a 95% confidence interval of (`r round(disp_slope_ci[1], 3)`, `r round(disp_slope_ci[2], 3)`).


```{r}
#| label: fig-disp
#| fig-width: 4
#| fig-height: 3
#| fig-cap: |
#|     Relationship across simulation runs between relative dispersal rate
#|     and speciation. Points are jittered around 1 and 0 and are color coded 
#|     according to local density of points all in an attempt to more clearly 
#|     show the distribution of simulation results. Curve is a 
#|     binomial generalize linear model and gray region shows 95% confidence 
#|     envelope. 


logistic_hist_plot(sim_dat, (1 + m_prop) * g / la, xlog = TRUE, 
          xlab = "Relative dispersal rate", 
          ymin = -0.003, ymax = 0.02,
          formula = speciation ~ log((1 + m_prop) * g / la))
```

We might also expect that speciation decreases with the duration of the wait time $\tau$ to full speciation. However, $\tau$ is only the initial value for the wait time, dispersal continually augments this wait time, thus $\tau$ by itself does not impact probability of speciation.

```{r}
#| label: tau-mod
#| message: false

# model of speciation changing with wait time to full speciation
tau_mod <- glm(speciation ~ log(tau), 
                data = sim_dat, family = binomial)
tau_slope_ci <- confint(tau_mod, 2)
```

The 95% confidence interval for the slope of speciation probability against $\log(\tau)$ is (`r round(tau_slope_ci[1], 3)`, `r round(tau_slope_ci[2], 3)`), which does skew toward negative values, but includes the value of slope = 0.


### Evaluating the relationship between abundance and speciation

These results are reported in the main text. Here we show how the results from the main text are derived. 

First we build a quadratic model of probability of speciation as a function of abundance with the below code:

```{r}
#| label: sim-mod
#| cache: true
#| dependson: run-sim

sim_dat$log_abund <- log(sim_dat$mean_pop_size)
sim_mod <- glm(speciation ~ log_abund + I(log_abund^2), 
               family = binomial, data = sim_dat)

sim_quad_ci <- confint(sim_mod, parm = 3)

```


The confidence interval for the quadratic term is far from 0:

```{r}
sim_quad_ci
```

We now turn to developing the figure of abundance versus speciation. These results are reported in the main text, but the code below is used to create the visualization of speciation versus abundance.

```{r}
#| label: make-sim-fig
#| cache: true
#| dependson: run-sim


sim_fig <- logistic_hist_plot(sim_dat, mean_pop_size, xlog = TRUE, 
          xlab = "Mean population size", 
          ymin = 0, ymax = 0.018,
          formula = speciation ~ log(mean_pop_size) + I(log(mean_pop_size)^2))

sim_fig

# special_log10_breaks <- function(x) {
#     p <- seq(floor(x[1]),
#              ceiling(x[2]),
#              by = (floor(x[2]) - ceiling(x[1])) / 15)
#     
#     10^p
# }
# 
# spec_by_bin <- mutate(sim_dat, 
#                       bins = cut(
#                           mean_pop_size, 
#                           log(mean_pop_size, 10) |> 
#                               range() |> 
#                               special_log10_breaks() 
#                       )) |> 
#     summarize(spec_prop = mean(speciation), 
#               n = n(),
#               .by = bins) |> 
#     mutate(xlo = gsub("\\(|,.*", "", bins) |> as.numeric(), 
#            xhi = gsub("\\]|.*,", "", bins) |> as.numeric()) |> 
#     arrange(xlo)
# 
# # plotting
# sim_fig <- spec_plot(sim_dat, mean_pop_size, xlab = "Mean population size", 
#                xlog = TRUE, formula = y ~ x + I(x^2), 
#                se = FALSE) + 
#     geom_rect(data = spec_by_bin,
#               mapping = aes(xmin = xlo, xmax = xhi,
#                             ymin = 0, ymax = spec_prop),
#               alpha = 0.25,
#               inherit.aes = FALSE) 
# 
# # change order of layer display
# sim_fig$layers <- sim_fig$layers[c(1, 3, 2)]
# 
# # add background bars
# sim_fig <- sim_fig + 
#     theme(panel.background = element_rect(fill = "gray90"),
#           panel.grid.major.x = element_line(color = "white"),
#           panel.grid.minor.x = element_line(color = "white")) +
#     scale_x_log10(minor_breaks = spec_by_bin$xlo, expand = FALSE)
```

## Analyzing abundance and richness of arthropod genera in Hawai\`i {#sec-bgeom}

The results of analyzing Hawai\`i arthropod data are presented in the main text. The code to produce these results is shown here. 


### Preparing the data

First we load and prepare the data. Preparation involves filtering data to only endemic taxa and compiling average per-species abundance for genera based on data from @gruner2007 and compiling species richness for genera based on the Bernice Pauahi Bishop Museum arthropod checklist [@hichecklist].

```{r}
#| label: arth-data
#| cache: true

# read gruner data
arth <- read.csv("raw_data/arth.csv")

# read checklist data
arth_check <- read.csv("raw_data/hawaii_arthropod_checklist.csv")

hmean <- function(x) length(x) / sum(1 / x)

# calculate abundances of genera from survey data
gen_abund <- filter(arth, origin == "native-endemic") |> 
    group_by(genus, species_code) |>
    summarize(abund = n()) |>
    ungroup(species_code) |>
    summarize(mean_abund = median(abund), 
              survey_nspp = n_distinct(species_code)) |>
    ungroup()


# calculate richness of genera from checklist
gen_rich <- filter(arth_check, Genus != "") |>
    group_by(Genus) |>
    summarize(nspp = n_distinct(Genus.and.species), 
              status = first(Status)) |>
    ungroup() |> 
    rename(genus = Genus) |> 
    # make `status` column indicate if genus is found in survey data
    mutate(status = ifelse(genus %in% gen_abund$genus, 
                           "in survey", 
                           status))

# `gen` holds data ready to analyze for richness ~ abundance
gen <- inner_join(gen_rich, gen_abund)
```


### Model of within genus richness and average per-species abundance {#sec-arth-mod}

With the prepared data we can compute a quadratic and linear model of genus richness versus abundance. As explained in the main text, these models are beta-binomial generalized linear models and are achieved with custom functions available in the *abundolism* package associated with this paper. The main function is `glm_bgeom` which maximizes the log likelihood function and shares some common methods with `lm` and `glm` including `logLik` which enables likelihood analyses such as the likelihood ratio test. 

To summarize from the main text, our full quadratic model is 

$$
\begin{aligned}
y_i &\sim \text{BetaGeometric}(\alpha_i, \beta) \\
\log(\alpha_i) &= a_0 + a_1 x_i + a_2 x_i^2 \\
\log(\beta) &= b_0
\end{aligned}
$$

where $x_i$ is mean abundance of lineage $i$ and our estimate for the observable diversification probability for lineage $i$ is $1 - \hat{p_i} = 1 - \frac{\alpha_i}{\alpha_i + \beta}$.


```{r}
#| label: arth-mod
#| cache: true

# add column for log abundance
gen$log_abund <- log(gen$mean_abund)

# quadratic model
arth_mod <- glm_bgeom(nspp ~ log_abund + I(log_abund^2), data = gen)

# straight line model
arth_mod0 <- glm_bgeom(nspp ~ log_abund, data = gen)

# view the model output
arth_mod
arth_mod0
```

The likelihood ratio test comparing quadratic and linear models is achieved with a custom function `lrt_bgeom` from the *abundolism* package. 

```{r}
#| label: lrt
#| cache: true
#| dependson: [arth-data, arth-mod]

arth_lrt <- lrt_bgeom(arth_mod0, arth_mod)
```


To produce visualizations of the arthropod data we make use of a custom `ci_bgeom` function from *abundolism* that calculates confidence intervals and predictions from the output of `glm_bgeom`. This function first calculates standard errors for the $\log(\alpha)$ and $\log(\beta)$ terms of the beta-geometric GLM using the Wald approximation and then computes the standard error for $1 - \hat{p}$ using a change of variable approach. The predicted value $\hat{y}$ for numbers of species per genus is then $\hat{y} = \frac{1}{1 - \hat{p}}$ and its standard error is calculated from that of $1 - \hat{p}$ again with a change of variable approach.

```{r}
#| label: make-fig-arth
#| fig-width: 6
#| fig-height: 4
#| cache: true
#| dependson: [arth-data, arth-mod]

arth_mod_ci <- cbind(mean_abund = gen$mean_abund, 
                     ci_bgeom(arth_mod, type = "response"))

arth_p_ci <- cbind(mean_abund = gen$mean_abund, 
                   ci_bgeom(arth_mod, type = "prob"))

gp_nspp <- ggplot(arth_mod_ci, aes(mean_abund, ymin = y_lwr, ymax = y_upr)) +
    geom_ribbon(fill = "gray60", alpha = 0.5) +
    scale_x_log10() +
    scale_y_log10() +
    xlab("Mean genus abundance") +
    ylab("Genus richness") +
    geom_pointdensity(data = gen, mapping = aes(mean_abund, nspp), 
                      inherit.aes = FALSE) +
    scale_color_viridis_c() +
    geom_line(data = arth_mod_ci, 
              mapping = aes(mean_abund, y_hat), 
              linewidth = 1) +
    theme_cowplot()


gp_prb <- ggplot(arth_p_ci, aes(mean_abund, y = p_hat, 
                                ymin = p_lwr, ymax = p_upr)) +
    geom_ribbon(fill = "gray60", alpha = 0.5) +
    geom_line(linewidth = 1) +
    scale_x_log10() +
    ylim(0.5, 1) +
    xlab("Mean genus abundance") +
    ylab(expression(italic(1 - hat(p)))) +
    theme_cowplot()

gp_arth <- plot_grid(gp_prb, gp_nspp, rel_widths = c(0.75, 1), 
                     labels = c("A", "B"), label_x = 0.02)
```

### Evaluating impact of outliers {#sec-outlier}

There is one clear outlier: *Drosophila* with high richness given abundance. To make sure this outlier does not bias results we re-run the analysis of genus richness versus abundance without *Drosophila*.



```{r}
arth_nodro <- glm_bgeom(nspp ~ log_abund + I(log_abund^2), 
                      data = gen[gen$nspp < 400, ])
arth_nodro0 <- glm_bgeom(nspp ~ log_abund, 
                       data = gen[gen$nspp < 400, ])
```

Based on a likelihood ratio test we find the quadratic model is still preferred.

```{r}
lrt_bgeom(arth_nodro0, arth_nodro)
```

### Evaluating impact of missing genera

In merging the checklist data (providing species richness) with field survey data (providing abundance estimates) some genera were lost due to not being detected in the survey data.

```{r}
# endemic checklist genera not found in danʻs data
# only need to search for "end" because those with
# status "end" that are in gruner data had their
# status switched to "in survey"
n_checklist_missing <- sum(gen_rich$status == "end")
n_checklist_missing

```

`r n_checklist_missing` endemic genera from the checklist were absent from the survey data, representing `r round(100 * n_checklist_missing / sum(gen_rich$status %in% c("end", "in survey")), 2)`% of endemic genera from the checklist.

While this is a large percentage, it is also not unexpected. The survey data constitute a very small sample of the geographic space of the Hawaiian Archipelago as well as a small sample of the extremely heterogeneous environmental space of the archipelago. It is still worth evaluating how the distributions of species richnesses differ between those genera included in the survey data and those missing. @fig-rich-comp shows that genera not found in the survey data are highly skewed toward those with low species richness.

```{r}
#| label: fig-rich-comp

filter(gen_rich, status %in% c("end", "in survey")) |>
    mutate(status = factor(ifelse(status == "end", "no", "yes"),
                          levels = c("yes", "no"))) |>
    ggplot(aes(status, nspp)) +
    geom_violin(fill = "gray") +
    scale_y_log10() +
    theme_cowplot() +
    xlab("Found in survey data") +
    ylab("Number of species in genus")
```

Any evaluation of how these missing genera might change the relationship of average per-species abundance and genus richness (estimated in @sec-arth-mod) will be compromised because it is impossible to know the abundances of the species in the missing genera. These genera could be absent from the survey data because their constituent species are rare (i.e. low average per-species abundance) or because their constituent species are restricted to different habitats and/or islands (i.e. potentially high average per-species abundance found elsewhere). However, for the sake of producing any kind of evaluation, we assume that if any of these missing genera were to be present in the survey data, they would be very rare. Thus, we re-run our analysis of the relationship between average per-species abundance and genus richness with artificial data (representing these missing genera) augmented onto the real data. We augment the data with the real species richness of each missing genus and an average per-species abundance of 1. The analysis of these augmented data yields the same answer: the hump-shaped relationship between average per-species abundance and genus richness is strongly supported by a likelihood ratio test

```{r}
#| label: what-if

# what if we had a bunch of rare, low richness genera
gen_what_if <- rbind(select(gen, c(genus, nspp, log_abund)),
                     data.frame(genus = paste0("fake_", 1:n_checklist_missing),
                                # actual abundances of missing genera
                                nspp = 1, #gen_rich$nspp[gen_rich$status == "end"],
                                # log(1) = 0
                                log_abund = 0))

# quadratic model
what_if_mod <- glm_bgeom(nspp ~ log_abund + I(log_abund^2), data = gen_what_if)

# straight line model
what_if_mod0 <- glm_bgeom(nspp ~ log_abund, data = gen_what_if,
                          a_init = c(1, 1))

lrt_bgeom(what_if_mod0, what_if_mod)
```

### Assessing the choice of average per-species abundance {#sec-tot-abund}

We should evaluate if the choice of mean per-species abundance produces any unintended artifacts in our analysis of how mean per-species abundance relates to within genus richness. The primary concern would be if genera themselves somehow had a carrying capacity, then the more species contained in a genus, the less abundant, on average, each species would have to be. Two observations from the data dispel this concern: first, the vast majority of genera are represented by only a single species at a sampling location [@fig-site-gen-rich], making the partitioning of a hypothetical genus-level carrying capacity moot at all but the few sites where genera are represented by multiple species. 

```{r}
#| label: fig-gen-site-rich
#| fig-width: 4
#| fig-height: 3
#| fig-cap: |
#|     Distribution of numbers of species per genus observed in survey data 
#|     [@gruner2007].

# computing number of species per genus and total genus abundance per site
gen_by_site <- summarize(arth, 
          rich = n_distinct(species_code),
          tot_abund = n(),
          .by = c(site, genus))

ggplot(gen_by_site, aes(rich)) +
    geom_histogram(binwidth = 1, color = "black", fill = "gray") +
    scale_x_continuous(breaks = 1:12) +
    theme_cowplot()

```

Next, if a hypothetical genus carrying capacity were partitioned across multiple species, then at sites with multiple congeners there should be a flat relationship between number of congeners and total genus abundance per site. However, @fig-gen-rich-trend shows there is a positive relationship between number of congeners and total genus abundance, suggesting that the abundance of any one species in a genus does not trade-off with the abundances of other congeners, but rather they all add to a larger total genus abundance. The positive relationship is significant by a likelihood ratio test applied to a Poisson GLM of abundance predicted by log number of congeners (note: excluded are any data points in which only a single species in a genus is present at a site).

```{r}
#| label: gen-rich-lrt

gen_site_mod <- glm(tot_abund ~ log(rich), 
                    data = gen_by_site[gen_by_site$rich > 1, ], 
                    family = "poisson")

anova(gen_site_mod)
```

```{r}
#| label: fig-gen-rich-trend
#| fig-width: 4
#| fig-height: 3
#| fig-cap: |
#|     Relationship between number of congeners at a site and total genus 
#|     abundance at that site. Trend line (black line) and 95% confidence envelope 
#|     (gray shaded region) are from a Poisson GLM. Data p


    
ggplot(gen_by_site[gen_by_site$rich > 1, ], 
       aes(rich, tot_abund)) +
    geom_smooth(color = "black", 
                method = "glm", 
                method.args = list(family = "poisson")) +
    geom_pointdensity() +
    scale_x_log10() +
    scale_y_log10() +
    scale_color_viridis_c()

```

## References
