---
title: "Supplement to: *Intermediate abundance promotes speciation when dispersal is limited*"
format:
  pdf:
    number-sections: true
    keep-tex: true
    include-in-header: 
      text: |
        \renewcommand{\thefigure}{S\arabic{figure}}
        \renewcommand{\thesection}{S\arabic{section}}
crossref:
  fig-title: "Supplementary Figure"
  fig-prefix: "Supplementary Figure"
csl: oikos.csl
bibliography: abundolism.bib
---


## Computation set-up

We created a make use of a custom R package *abundolism* which can be installed from github. 

```{r}
#| label: ab-setup
#| eval: false

devtools::install_github("ajrominger/abundolism")
```

The following R computing environment is used for all simulations and analyses:

```{r}
#| label: setup

library(abundolism)
library(dplyr)
library(ggplot2)
library(ggpointdensity)
library(cowplot)
library(knitr)

sessionInfo() |> 
    print(locale = FALSE)

```

The quarto [@quarto] document used to generate this supplement can be found with the R command `system.file("ab_supp.qmd", package = "abundolism")`.


## Simulation details {#sec-sim}

The simulation tracks

- all local population sizes
- running mean population size per population, per iteration 
- time relative to all rates (i.e. not simply number of iterations)
- whether an incipient speciation event has occurred
- the wait time to full speciation
- whether full speciation has occurred

Each iteration an event is chosen at random. Possible events for each local population are

- local birth
- local death
- local speciation
- local immigration

There is also one possible global event: immigration from the source pool.

If there are $n_p$ local populations then the total number of possible events are $n_p \times 4 + 1$.

Given that the size of a local population is written $x_i$, the probability of each event is proportional to its respective rate:


\begin{align}
\mathbb{P}(\text{birth in pop. } i) &= \frac{\lambda x_i}{R} \\
\mathbb{P}(\text{death in pop. } i) &= \frac{\mu x_i}{R} \\
\mathbb{P}(\text{speciation in pop. } i) &= \frac{\nu x_i}{R} \\
\mathbb{P}(\text{local immigration from pop. } i) &= \frac{m_{prop}\gamma x_i}{R} \\
\mathbb{P}(\text{global immigration}) &= \frac{\gamma}{R}
\end{align}


where $R = \gamma + \left( \sum_i x_i \right) (\lambda + \mu + \nu + m_{prop} \gamma)$ is a normalization constant.

In the case of immigration (both local and from the source pool) the recipient of the dispersing individual must also be chosen. We choose the recipient local population at random with equal probability across local populations.

Once an event is randomly sampled, the variables being tracked are updated. Birth and immigration (local or global) augment a local population by 1. Death decrements a local population by 1. Immigration events have no impact on the source or the dispersing individual. Time is updated following the Gillespie algorithm: the sojourn time between events is a random variable with an exponential probability distribution with rate R: $s \sim Exp(R)$. Each time step a random sojourn time is sampled and adding to the running total time since the start of the simulation.

If a population experiences local extirpation all of its speciation-related information is re-set including removing any incipient speciation and restoring the wait time to full speciation back to the default of $\tau$.

If full speciation is achieved the simulation stops. If no speciation occurs but the specified number of simulation iterations is reached, the simulation stops. Once stopped, the simulation returns the tracked variables of time, mean population size, and whether full speciation has occurred.



## Simulation experiment details


Simulation parameters are drawn from uniform distributions with the following minimum and maximum limits:

```{r}
#| label: par-range
#| cache: true

par_range <- data.frame(
    param = c("la", "mu", "g", "m_prop", "nu", "tau", "xi"), 
    min   = c(0.1,  0.1,  0.1,  0,        0,     20,  10), 
    max   = c(4,    4,    1,    0.1,      0.1,  200, 100)
)


# fixed params
np <- 10
nstep <- 10000

# number of simulation replicates
nrep <- 5000

# format param names for printing
mutate(par_range, 
       param = sprintf("`%s`", param)) |> 
    kable(format.args = list(scientific = FALSE))
```

The parameters governing number of populations (`np`) and number of time steps (`nstep`) do not vary across simulations and are set to `np =` `r np` and `nstep =` `r nstep`. We simulate a total of `nrep =` `r nrep` replicates, each with a unique, randomly sampled set of parameter values.


### Running the simulation experiment

Parameter values for each simulation replicate are stored in a `data.frame` with one column for each parameter.

```{r}
#| label: par-sample
#| cache: true

# change or remove for different results
set.seed(123)

# data.frame of randomly sampled param values
# one param per column
pars <- sapply(par_range$param, function(p) {
    runif(nrep, 
          par_range[par_range$param == p, "min"], 
          par_range[par_range$param == p, "max"])
}) |> 
    as.data.frame()

```

With those parameters sampled, we can now run the simulation, one realization for each parameter combination. For details on inputs and outputs of the simulation function `sim_BDI_spec` see the help documentation accessible with `?sim_BDI_spec`.

```{r}
#| label: run-sim
#| cache: true
#| dependson: [par-range, par-sample]

# now run the simulation
sim_dat <- sim_BDI_spec(la = pars$la, mu = pars$mu, g = pars$g, 
                        m_prop = pars$m_prop, nu = pars$nu, 
                        tau = pars$tau, xi = pars$xi, 
                        np = np, nstep = nstep)

```

### Interpreting the simulation experiment

Now we can visually inspect the results. Our main purpose is to study how abundance relates to speciation in this model, but first we check other relationships to ensure the model behaves as expected. In particular, population size should be larger when $\lambda > \mu$ and smaller, supported primarily by immigration. When $\lambda < \mu$ populations will grow indefinitely. When birth and death rates are equal ($\lambda = \mu$) the population could crash or expand depending on small random fluctuations and random immigration. @fig-la-mu shows exactly this pattern. Of note: simulation runs experiencing speciation diverge from the expected pattern but that is specifically because our simulation algorithm stops the simulation short (i.e. populations are not allowed to continue changing) once speciation occurs.

```{r}
#| label: fig-la-mu
#| eval: false
#| fig-width: 4
#| fig-height: 3

ggplot(sim_dat, aes(la - mu, mean_pop_size)) +
    geom_point(data = select(sim_dat, !speciation), 
               mapping = aes(la - mu, mean_pop_size), color = "gray50") +
    geom_pointdensity() + 
    facet_wrap(vars(speciation), 
               labeller = as_labeller(c(`0` = "No speciation", 
                                        `1` = "Yes speciation"))) +
    scale_y_log10() + 
    scale_color_viridis_c(trans = "log10")


```

```{r}
#| label: nu-mod
#| message: false

# model of speciation changing with nu
nu_mod <- glm(speciation ~ nu, data = sim_dat, family = binomial)
nu_slope_ci <- confint(nu_mod, 2)


```




We also expect that probability of specification should increase with the parameter $\nu$, the rate of incipient speciation. @fig-nu confirms this expectation but shows the relationship to be weak, indicating that while this parameter obviously matters, its effect is modulated by other factors such as abundance and dispersal. A logistic regression model confirms this weakly positive but meaningful relationship estimating a 95% confidence interval on the slope of (`r round(nu_slope_ci[1], 3)`, `r round(nu_slope_ci[2], 3)`).


```{r}
#| label: fig-nu
#| fig-width: 4
#| fig-height: 3

spec_plot(sim_dat, nu, 
          xlab = expression("Speciation rate"~italic(nu)))

```


```{r}
#| label: disp-mod
#| message: false

# model of speciation changing with dispersal
disp_mod <- glm(speciation ~ I(log((g + m_prop * g) / la)), 
                data = sim_dat, family = binomial)
disp_slope_ci <- confint(disp_mod)
```

Additionally, we expect dispersal to decrease the chance of speciation because dispersal prolongs the wait time until full speciation. But we have to evaluate this expectation with respect to *relative* dispersal, that is dispersal relative to birth because these rate trade-off in determining whether a population grows due to intrinsic (birth) or extrinsic (dispersal) forces. Dispersal comes from two sources: global dispersal and local dispersal, so we add these two sources when considering total dispersal. @fig-disp Confirms the negative relationship between speciation and dispersal. The slope from a logistic regression model for log dispersal normalized by birth rate has a 95% confidence interval of (`r round(nu_slope_ci[1], 3)`, `r round(nu_slope_ci[2], 3)`).


```{r}
#| label: fig-disp
#| fig-width: 4
#| fig-height: 3

spec_plot(sim_dat, (1 + m_prop) * g / la, xlog = TRUE, 
          xlab = "Relative dpeciation rate")

spec_plot(sim_dat[sim_dat$g > 0.5, ], tau, xlog = TRUE, 
          xlab = "Wait time")
```


```{r}
#| label: tau-mod
#| message: false

# model of speciation changing with wait time to full speciation
tau_mod <- glm(speciation ~ log(tau) * I(log((1 + m_prop) * g / la)), 
                data = sim_dat, family = binomial)
confint(tau_mod)
```

We might also expect that speciation decreases with the duration of the wait time $\tau$ to full specaition. However, $\tau$ is only the initial value for the wait time, dispersal continually augments this wait time, thus $\tau$ by itself does not impact probability of speciation.


### Evaluating the relationship between abundance and speciation

These results are reported in the main text. Here we show how the results from the main text are derived. 

First we build a quadratic model of probability of speciation as a function of abundance with the below code:

```{r}
#| label: sim-mod
#| cache: true
#| dependson: run-sim

sim_dat$log_abund <- log(sim_dat$mean_pop_size)
sim_mod <- glm(speciation ~ log_abund + I(log_abund^2), family = binomial, data = sim_dat)

sim_quad_ci <- confint(sim_mod, parm = 3)

```


The confidence interval for the quadratic term is far from 0:

```{r}
sim_quad_ci
```

We now turn to developing the figure of abundance versus speciation. 

```{r}
#| label: make-sim-fig
#| cache: true
#| dependson: run-sim

sim_fig <- spec_plot(sim_dat, mean_pop_size, xlab = "Mean population size", 
          xlog = TRUE, formula = y ~ x + I(x^2))

```

## Analyzing abundance and richness of arthropod genera in Hawai\`i

The results of analyzing Hawai\`i arthropod data are presented in the main text. The code to produce these results is shown here. First we load and prepare the data. Preparation involves compiling mean abundance for genera based on data from @gruner2007 and compiling species richness for general based on the Bernice Pauahi Bishop Museum arthropod checklist [@hichecklist].

```{r}
#| label: arth-data
#| cache: true

# consider making exported data and sourcing from package
# read gruner data
arth <- read.csv("raw_data/arth.csv")

# read checklist data
arth_check <- read.csv("raw_data/hawaii_arthropod_checklist.csv")

# calculate richness of genera from checklist
gen_rich <- filter(arth_check, Genus != "") |>
    group_by(Genus) |>
    summarize(nspp = n_distinct(Genus.and.species)) |>
    ungroup() |> 
    rename(genus = Genus)

# calculate abundances of genera from survey data
gen_abund <- group_by(arth, genus, species_code) |>
    summarize(abund = n()) |>
    ungroup(species_code) |>
    summarize(mean_abund = mean(abund)) |> 
    ungroup()


# `gen` holds data ready to analyze for richness ~ abundance
gen <- inner_join(gen_rich, gen_abund)
```

With the prepared data we can compute a quadratic and linear model of genus richness versus abundance. As explained in the main text, these models are beta-binomial generalized linear models and are achieved with functions available in the *abundolism* package [@abundolism-package] associated with this paper. The main function is `glm_bgeom` which shares some common methods with `lm` and `glm` including `logLik`. 


```{r}
#| label: arth-mod
#| cache: true

# add column for log abundance
gen$log_abund <- log(gen$mean_abund)

# quadratic model
arth_mod <- glm_bgeom(nspp ~ log_abund + I(log_abund^2), data = gen)

# straight line model
arth_mod0 <- glm_bgeom(nspp ~ log_abund, data = gen)

# view the model output
arth_mod
arth_mod0
```

The likelihood ratio test comparing quadratic and linear models is achieved with a custom function `lrt_bgeom` from the *abundolism* package. 

```{r}
#| label: lrt
#| cache: true
#| dependson: [arth-data, arth-mod]

arth_lrt <- lrt_bgeom(arth_mod0, arth_mod)
```


To produce visualizations of the arthropod data we make use of a custom `ci_bgeom` function from *abundolism* that calculates confidence intervals and predictions from the output of `glm_bgeom`.

```{r}
#| label: fig-arth
#| fig-width: 6
#| fig-height: 4
#| cache: true
#| dependson: [arth-data, arth-mod]

arth_mod_ci <- cbind(mean_abund = gen$mean_abund, 
                     ci_bgeom(arth_mod, type = "response"))

arth_p_ci <- cbind(mean_abund = gen$mean_abund, 
                   ci_bgeom(arth_mod, type = "prob"))

gp_nspp <- ggplot(arth_mod_ci, aes(mean_abund, ymin = y_lwr, ymax = y_upr)) +
    geom_ribbon(fill = "gray") +
    scale_x_log10() +
    scale_y_log10() +
    xlab("Mean genus abundance") +
    ylab("Genus richness") +
    geom_pointdensity(data = gen, mapping = aes(mean_abund, nspp), 
                      inherit.aes = FALSE) +
    scale_color_viridis_c() +
    geom_line(data = arth_mod_ci, 
              mapping = aes(mean_abund, y_hat), 
              linewidth = 1)


gp_prb <- ggplot(arth_p_ci, aes(mean_abund, y = p_hat, ymin = p_lwr, ymax = p_upr)) +
    geom_ribbon(fill = "gray") +
    geom_line(linewidth = 1) +
    scale_x_log10() +
    ylim(0, 1) +
    xlab("Mean genus abundance") +
    ylab(expression(italic(1 - hat(p))))

gp_arth <- plot_grid(gp_prb, gp_nspp, rel_widths = c(0.75, 1), 
                     labels = c("A", "B"), label_x = 0.02)
gp_arth
```


There are two outliers (*Drosophila* with high richness given abudnance, and *Salina* with highest overall abundance). To make sure these do not bias results we re-run analysis without them.



```{r}
arth_mod <- glm_bgeom(nspp ~ log_abund + I(log_abund^2), 
                      data = gen[gen$nspp < 400 & gen$mean_abund < 1000, ])
arth_mod0 <- glm_bgeom(nspp ~ log_abund, 
                       data = gen[gen$nspp < 400 & gen$mean_abund < 1000, ])
```

Based on a likelihood ratio test we find the quadratic model is still prefered

```{r}
lrt_bgeom(arth_mod0, arth_mod)
```


## References
