---
title: "Intermediate abundance promotes speciation when dispersal is limited"
format:
  pdf:
    keep-tex: true
    linenumbers: true
    include-in-header:
      text: |
        \usepackage{authblk}
        \usepackage{orcidlink}
        \usepackage{xr}
        \usepackage{lineno}
        \renewcommand{\sfdefault}{\rmdefault}
        \externaldocument{ab_supp}
        \newcommand{\sfig}[1]{Supplementary Figure~\ref{#1}}
        \newcommand{\ssec}[1]{Supplementary Section~\ref{#1}}
csl: oikos.csl
bibliography: abundolism.bib
---

\linenumbers

## Abstract 


Why do some lineages diversify while others do not? This remains a central question in evolutionary ecology. A long-standing assumption, dating to Darwin and embedded in the Unified Neutral Theory of Biodiversity, holds that abundant species should speciate at higher rates. Conversely, theoretical and empirical work highlights the possibility that rare and dispersal-limited clades might be more prone to speciation. Using a birth-death-immigration model with protracted speciation in a multi-population landscape connected by limited dispersal, we show that abundance has a hump-shaped effect on probability of speciation. Our model reveals that intermediate abundance maximizes speciation probability because larger populations disperse more, swamping regional differentiation and inhibiting speciation completion, while smaller populations lack the persistence and incipient speciation needed to diversify. We find empirical support for this prediction with an analysis of data from arthropods endemic to Hawai\`i, where genus-level species richness shows a significant hump-shaped relationship with mean genus abundance. These findings provide a mechanistic explanation for a nuanced relationship between abundance and diversification.

\pagebreak

## Introduction


```{r}
#| label: setup
#| include: false

library(knitr)
library(kableExtra)
library(dplyr)
library(ggplot2)
library(ggpointdensity)
library(grid)
library(sf)

opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

# helper function to load cache from supp
load_supp <- function(cname) {
    f <- sprintf("%s.*RData", cname) |> 
        grep(list.files("ab_supp_cache/pdf", full.names = TRUE), 
              value = TRUE) |> 
        gsub("\\.RData", "", x = _)
    
    o <- lazyLoad(f, envir = globalenv())
}

```

```{r}
#| label: supp
#| include: false
#| cache: true
#| cache.extra: !expr file.info("ab_supp.qmd")$mtime

# render the supplement so can be referenced and used in main

quarto::quarto_render("ab_supp.qmd", quiet = FALSE)
```

```{r}
#| label: supp-tex-render
#| include: false

# need .aux file for cross referencing
tinytex::xelatex("ab_supp.tex", clean = FALSE)
```

A fundamental question in evolutionary ecology concerns why some clades are diverse while others are not. Are there properties of species and lineages that promote speciation? One line of inquiry with a long and convoluted history is the role of commonness and rarity in the process of diversification. Going back over 150 years, @darwin argued that widespread, abundant species should be superior competitors and thus more likely to give rise to new species. 

Building on this foundation, Yule's -@yule model, which became a landmark contribution to modeling diversification, was initially intended as a way to incorporate demography into predictions for diversification and clade richness [@pennell-mac2025]. Despite these deep roots in evolutionary thinking, little consensus has emerged on the relationship between abundance and diversification. Darwin's line of thinking carried forward to macroecologists who attempted to model, with some empirical support, the presumed positive association between commonness, fitness, and speciation [@maurer1999]. 

In parallel, paleontologists searched for relationships between commonness and diversification, alternately finding evidence for a positive relationship [@krug2008] but also a negative relationship [@stanley1986; @jablonski2003]. Empirical studies of modern taxa have largely found that commonness is negatively or inconclusively related to measures of diversification [@jablonski2003; @makarieva2004; @greenberg2017; @smyvcka2023; @afonso2025; but see @hay2022]. While some of these studies focus specifically on average abundance, others on range size, and some on both, the two macroecological properties are most often strongly correlated, a fascinating macroecology pattern in and of itself [@brown1995; @gaston2002]. This correlation may be caused, at least partly, by the net increase in dispersal of larger populations owing simply to chance: if each individual carries a probability of dispersing, more individuals in a population will lead to more overall dispersal [@hubbell2001]. It is also of course possible that shared underlying mechanisms cause both increased population growth rate and increased dispersal rate [@gaston1997; @holt1997].

More recent theoretical and modeling work has largely assumed a positive relationship between commonness and diversification. @hubbell2001 perhaps started this trend with the inclusion of “point mutation” speciation in the Unified Neutral Theory of Biodiversity (UNTB). Because point mutation speciation assumes a constant probability of speciation for every individual, lineages with more individuals will experience more speciation events in the UNTB [@hubbell2001; @etienne2007modes]. The speciation mechanism first proposed by @hubbell2001 was never an accurate model of real diversification [@ricklefs2003], but even later attempts to increase the realism of speciation in the UNTB retain the emergent property that more abundant lineages will undergo more speciation. Protracted speciation [@rosindell2010protracted] again assumes that incipient speciation is constant across individuals and thus full speciation will scale positively with lineage abundance; fission speciation [@etienne2011fission] also assumes the probability of fission (the event that leads to speciation) increases with the number of individuals in a species.  Largely independently from UNTB and its descendants, phylogenetic models of geograpsic change and diversification also explicitly assume that more widespread species have greater opportunity for speciation [@goldberg2011].

The process that underlies speciation itself may suggest something different in terms  of the relationship between speciation, commonness, and rarity: dispersal.  Dispersal is a key mechanism by which populations can become isolated or connected, potentially leading to speciation or admixture [@yamaguchi2022].  Dispersal figures prominently in phylogenetic models of the geograpsic range and diversification [@matzke2014], it is necessary to maintain local biodiversity in the UNTB [@hubbell2001], and it is the mechanism connecting abundance to range size in both neutral [@hubbell2001] and non-neutral [@brown1995] macroecological models. 

Due to the potential for dispersal to both create population isolates but also swamp out regional differences that could have led to speciation, the connection between dispersal ability as a biological trait and speciation as an evolutionary outcome remains a fertile line of inquiry. Some empirical studies have found that measures of diversification have a hump-shaped or negative relationship with morphological proxies for dispersal ability [@price2004; @claramunt2012; @czekanski2019] while others have found a positive to flat relationship between dispersal ability and diversification [@claramunt2025]. Agent-based model simulations have supported the idea of a trade-off between too little dispersal leading to population instability and too much dispersal leading to admixture with intermediate dispersal balancing the two extremes allowing for speciation [@birand2012; @ashby2020; @ciccheto2024]. 

But dispersal is not just a trait connected to morphology or imposed in a simulation, it emerges from population dynamics: each individual carries some per-capita probability of dispersing, but because that probability is applied across every individual in the population, the total number of dispersal events scales linearly with abundance, thus species with more individuals produce more dispersal overall. This is the assumption, a realistic one, in birth-death-immigration models [@kendall1948]. Therefore, if commonness and dispersal affect speciation, their effects are interconnected because population-level dispersal rate increases both with intrinsic per-capita dispersal probability and with population size. It also remains an open debate whether their effects result from the determinism of adaptive evolution as @darwin believed or instead emerge from chance.

Here we develop a birth-death-immigration model with protracted speciation embedded in a landscape of multiple local populations connected with limited dispersal to investigate the role of abundance and stochastic dispersal in modulating the probability of speciation. Contrary to @darwin and UNTB [@hubbell2001; @etienne2007modes; @rosindell2010protracted; @etienne2011fission], we find that intermediate abundances lead to the greatest probability for speciation. Critically, this result does not depend in any way on whether populations are more or less adaptively fit, it only depends on a balance between large enough population sizes to accrue sufficient probability of incipient speciation but small enough population sizes to not lose regional differentiation due to increased dispersal. We also analyze real data on the species richness and abundance of arthropod communities in Hawai\`i finding empirical evidence for intermediate abundance promoting speciation. The isolation of Hawai`i, leading to tractable endemic radiations, and its known geologic age, ranging from still forming to 4–6 million years, make it an ideal study system for investigating the interplay of abundance and speciation.


## Methods

### Simulating a birth-death-immigration process with protracted speciation

We simulate a birth-death-immigration process with protracted speciation in a modified metapopulation [@hanski1998] setting. For simplicity, we consider only a single species (not multi-species) metapopulation. Constraining our simulations to a single species allows us to focus on the processes of interest---population dynamics and speciation---without entering into debates on multi-species dynamics [@hillerislambers2012; @rosindell2012]. 

In this simulation framework there are a number of local populations (minimum two are needed, we simulate 10) fully connected by local dispersal as well as a global source population connected to each local population by dispersal. The global source population is considered extremely large relative to local populations such that extinction of the global source population is unlikely on the time scale of local dynamics (we approximate this assumption by never allowing the source population to go extinct). Local populations start at size 0 and grow and shrink according to local births and deaths, dispersal between local populations, and dispersal from the source pool. 

Our simulation model is consistent with other simulation models derived from the UNTB [@hubbell2001; @rosindell2013; @rosindell2015], but unlike the multi-species perspective of those models, we concern ourselves only with a single species and whether that one species undergoes a speciation event. In this way we do not make the assumption of per capita neutrality across species, meaning our results could generalize to multi-species eco-evolutionary models based on niche theories [e.g. @tilman1982; @chesson2000; @letten2017]. We also relax the zero-sum and fixed time step approaches of most UNTB-like models. Instead, we initialize all patches as empty and let their numbers of individuals grow and shrink with birth, death, and dispersal. We also use a Gillespie algorithm [@gillespie1976; @gillespie1977] to simulate dynamics. In the Gillespie algorithm each discrete time step is an event chosen at random in proportion to the rate of occurrence of that event (i.e. not the fixed death-replacement cycle of the UNTB; @hubbell2001). Although events are sampled at discrete time steps, time itself is measured in actual units of time, not time steps, by sampling sojourn times between events from an exponential distribution with rate equal to the sum of all rates for all possible events.

#### Protracted speciation

In our simulation model a new species arises via a modified protracted speciation process [@rosindell2010protracted] similar to the model of @tao2021. Protracted speciation was first proposed by @rosindell2010protracted to try to bring more realism to the speciation process of the UNTB which originally modeled speciation as a point mutation-like process in which a single individual gets randomly flipped to a new species identity with some small probability $\nu$ [@hubbell2001]. Protracted speciation instead assumes that this point mutation-like process, again with probability $\nu$, only flips an individual to be an *incipient* species; if the lineage of that incipient species survives $\tau$ time steps, the process is said to reach full speciation but if the lineage dies out before $\tau$ time steps, no new species is formed [@rosindell2010protracted].

We extend the protracted speciation model to a multi-patch simulation landscape. In our model, after a population is founded through immigration it, the newly founded population, has a small probability of becoming an incipient new species. If an incipient speciation event occurs, and in the absence of subsequent immigration, the population of the incipient new species must persist $\tau$ time units for full speciation. If the population does not persist for $\tau$ time units, the incipient new species is lost without ever becoming a fully new species.

After an incipient speciation event, subsequent immigration events slow the progress toward full speciation. We model this by adding a small increment to $\tau$ equal to $\xi / n_i$ where $\xi$ is a parameter we can set and $n_i$ is the size of the population containing the incipient species. Due to $n_i$ in the denominator, this slowdown will be felt less by large populations and more by smaller populations in agreement with population genetic theory on the effect of immigration on population differentiation [@wright1931; @mills1996].

@tao2021 also developed a similar mutli-patch protracted speciation model, but ours is different in three ways:

1. @tao2021 assume every population founded by immigration is an incipient species; we assume that a new population founded through immigration remains part of the parent species unless an explicit incipient speciation event occurs.
2. In our Gillespie algorithm approach, time to full speciation is measured in actual time units, not discrete time steps.
3. @tao2021 assume that a fixed constant is added to $\tau$ for every immigration event


#### Simulation rules

The preceeding sections explained out simulation model, here we summarize the rules by which the simulation progresses:

1. All local populations are initialized as empty, to be filled by immigration.
1. Per capita birth, death, immigration, and incipient speciation all happen independently and are determined by their own respective rates. 
1. Immigration is modeled as the dispersal of a newly born individual, thus the population from which the individual emigrates is unchanged; this birth-then-emigrate mechanism is a widely used simplification in models inspired by the UNTB [@hubbell2001; @rosindell2013; @rosindell2015; @tao2021].
2. Speciation has 2 steps:
    i. After a population is founded through immigration, incipient speciation can change that population to an incipient new species with a per capita rate of $\nu$. Subsequent incipient speciation events in the same local population are ignored because the population is already an incipient new species. Multiple incipient species can occur across different patches, but not within the same patch.
    ii. If an incipient species survives long enough it becomes a new species; "long enough" is determined by a waiting time parameter $\tau$---the larger $\tau$, the longer the time frame for a fully new species to emerge.
3. Immigration between local communities and from the global source pool slows the progress toward full speciation; if an incipient species has to wait $\tau$ time units (in the absence of immigration) until it is a full species, each immigration event adds an increment to $\tau$ of $\xi / n_i$. 
3. If an individual of an incipient species disperses to a different population, it is subsumed within that population and its incipient status is lost.
4. If a local population goes extinct, the speciation tracking for that patch is reset.
5. The simulation is stopped under one of two conditions: i) if any incipient species becomes a fully diverged new species, or ii) if the simulation reaches the maximum designated number of iterations.

Because of the stopping conditions, each simulation replicate will either yield one and only one full speciation, or no full speciation events at all. 



{{< include fig-concept.qmd >}}


Further details of the simulation are provided in \ssec{sec-sim}. All simulation code is written in R [@rcore] using *Rcpp* [@rcpp] and *RcppArmadillo* [@rcpparma]. Underlying C++ and R source code is available in the R package *abundolism* accompanying this paper. \ssec{sec-setup} explains how to access this custom R package.

### Simulation experiment

```{r}
#| label: pars-info-setup


# loads `par_range`, `np`, `nstep`, `nrep`
load_supp("par-range")
```


@tbl-pars summarizes the simulation parameters in our model and the range of values we use in our simulation experiment. In all cases parameter values are drawn from a uniform distribution. A total of `r format(nrep, big.mark = ",", scientific = FALSE)` simulations were run, each with a unique set of randomly sampled parameters. Parameter values and the number of iterations were chosen with three considerations in mind: 

1. speciation should be possible to observe in a reasonable number of simulation iterations for computational practicality
2. population sizes should be within realistic ranges
3. timescales should be within reasonable bounds

In the context of the speciation-related parameter values we used, timescales should be interpreted in units of 10,000 years. Thus the maximum value of $\tau$ is `r format(par_range$max[par_range$param == "tau"] * 10000, big.mark = ",", scientific = FALSE)` years and the maximum rate of incipient speciation $\nu$ is `r par_range$max[par_range$param == "nu"]` per individual per 10,000 years, leading to rates of new species divergence roughly in line with published estimates [@hedges2015].


```{r}
#| label: tbl-pars
#| tbl-cap: |
#|     Parameters governing the simulation and their ranges over the simulation experiment. Mathematical parameter names are given first followed by their name in the simulation code. Note `np` and `nstep` were fixed throughout the course of the simulation experiment


par_info <- data.frame(param = c("np", "nstep"), 
                       min = c(np, nstep), 
                       max = c(np, nstep)) |> 
    (\(x) rbind(par_range, x))() 

par_desc <- c(
    "per capita local birth rate", 
    "per capita local death rate", 
    "dispersal rate from the global source pool to one local population", 
    "proportional dispersal rate between local populations relative to global rate", 
    "per capita incipient speciation rate", 
    "wait time to full speciation in the absence of dispersal", 
    "amount each dispersal event sets back the progression toward speciation", 
    "number of local populations", 
    "number of simulation iterations"
) |> 
    setNames(par_info$param)


par_tab <- select(par_info, param) |> 
    mutate(
        par_name = case_match(param, 
                              "la" ~ "$\\lambda$", 
                              "mu" ~ "$\\mu$", 
                              "g" ~ "$\\gamma$", 
                              "m_prop" ~ "$m_\\text{prop}$", 
                              "nu" ~ "$\\nu$", 
                              "tau" ~ "$\\tau$", 
                              "xi" ~ "$\\xi$", 
                              "np" ~ "$n_p$", 
                              "nstep" ~ "$n_\\text{step}$"), 
        par_code = sprintf("(\\texttt{%s})", param) |> 
            gsub("_", "\\\\_", x = _),
        description = par_desc[param]
    ) |> 
    left_join(par_info) |> 
    select(-param) |> 
    relocate(description, .after = max)
    

kable(par_tab, booktabs = TRUE, format = "latex",
      col.names = NULL, align = "rrrrl", escape = FALSE, 
      format.args = list(big.mark = ",", scientific = FALSE)) |>
    add_header_above(c("parameter" = 2, "range" = 2, "description" = 1), 
                     line = TRUE) |> 
    column_spec(5, width = "3in")

```

Parameter values were chosen as a balance between biological realism and computational efficiency: rates are sufficiently fast to allow for abundances to reach levels often found in survey data (e.g. the arthropod survey data we analyze here) within `r format(nstep, big.mark = ",")` time steps.


```{r}
load_supp("run-sim")


```

#### Analysis of simulation experiment results

Our primary focus is on the relationship between abundance and probability of speciation. We do this by comparing the presence or absence of a full speciation event to species abundance averaged over time and patches. However, we must account for the stopping conditions of our simulation being dependent on the very process---speciation---that we are analyzing. Simulation runs reaching full speciation will have fewer time steps than those reaching the pre-determined time step limit. We account for this by cutting each simulation run into a set of time step windows of fixed width (i.e. fixed number of time steps). We then analyze the presence or absence of speciation and average abundance within these non-overlapping windows of fixed width. We apply this approach to the analysis of speciation versus abundance, but also to auxiliary analyses of how speciation and abundance respond to model parameters. Those auxiliary analyses include visualizing average abundance across different birth and death rates and fitting models of the probability of speciation as a function of model parameters.

In fitting any models with these simulation results cut into fixed-width windows, we much account for the non-independence between windows from the same model run. We do this using generalized linear mixed effects models where simulation run is a random effect.

### Analysis of arthropod richness and abundance from Hawai\`i

```{r}
load_supp("arth-data")
load_supp("miss-gen")
```

To bring real data to bear on our insights from simulations, we analyze the species richness and average per-species abundance of arthropod genera in Hawai\`i. Endemic arthropods in Hawai\`i are a uniquely ideal study system because, although only an approximation, most arthropod genera only arrived in Hawai\`i once, meaning all endemic species of those genera represent *in situ* diversification [@cowie2008]. For the purpose of our analysis, we proceed with this approximation. The unique situation in Hawai\`i, that each arthropod genus arrived approximately only once on this geologically young archipelago, also avoids the problem that taxonomic ranks are temporally inconsistent at a global level [@avise2011].

To calculate species richness per genus we use the Bernice Pauahi Bishop Museum's checklist of arthropods in Hawai\`i [@hichecklist]. This checklist represents the current understanding of terrestrial arthropod species richness in Hawai\`i, though is inevitably incomplete for a place supporting rapid diversification. To estimate average per-species abundance for each genus we use an extensive survey of arthropods from across the Hawaiian Archipelago [@gruner2007]. The survey data from @gruner2007 comprise sample locations from `r unique(arth[, c("site", "tree")]) |> nrow()` trees across `r unique(arth$site) |> length()` sites. Sites are located on Hawai\`i Island, Moloka\`i, and Kaua\`i and follow the Long Substrate Age Gradient plots established by @vitousek1995 that are constrained to be within similar montane rainforest environments. 

Combining the checklist with survey data resulted in the retention of `r sum(gen_rich$status %in% c("end", "in survey")) - n_checklist_missing` endemic genera out of `r sum(gen_rich$status %in% c("end", "in survey"))` total from the checklist, representing a loss of `r round(100 * n_checklist_missing / sum(gen_rich$status %in% c("end", "in survey")), 2)`%. While this is a large percentage, it is also not unexpected. The survey data constitute a very small sample of the geographic space of the Hawaiian Archipelago as well as a small sample of the extremely heterogeneous environmental space of the archipelago. \ssec{sec-miss-gen} evaluates the impact of these missing genera on our analyses, finding that our results are robust to this subsampling of total genera.

Average per-species abundance was estimated for each genus as the mean of each species's abundance for all species belonging to that genus across all sample locations. We use average per-species abundance to represent the measure of commonness or rarity of a genus because speciation occurs at the species level, or lower, so abundance of higher taxonomic groups is not as relevant. We can also estimate the average abundance of a species with lower error than the total abundance of a genus. In \ssec{sec-tot-abund} we further study the choice of using average per-species abundance and find it to be a rigorous one.

We are interested in determining if the number of species per genus (a proxy for diversification) peaks at intermediate per-species abundances. To do this we fit a quadratic generalized linear model with species richness as response and average per-species abundance as explanatory variable. But the question remains, what probability distribution should we assume for number of species? @kendall1948 showed that a birth-death process will result in a geometric distribution of descendants. Conditional on lineage survival at time $t$ after initial establishment, we can parameterize the geometric probability mass function as 

\begin{equation}
f(y) = (1 - \psi(t))^{y-1}\psi(t)
\end{equation}

where $y$ is the number of species in a lineage. $\psi(t)$ is the probability that a lineage survives but we do not observe any bifurcation events by time $t$. Thus $1 - \psi(t)$ is the probability of the complement which is observing at least one bifurcation event leaving multiple descendants at time $t$.  We will call $1 - \psi(t)$ the "observable diversification probability." $\psi(t)$ depends on rates of speciation and extinction in addition to time $t$. 

The number of species in any given lineage $i$ extant at time $t$ will follow a geometric distribution with probability parameter $p_i = \psi_i(t_i)$. But we do not know $p_i$ because we do not know the arrival times of most arthropod lineages in Hawai\`i---thus we do not know $t_i$---and we do not know their rates of speciation and extinction. However, we can model our uncertainty in $p_i$ by letting it be a random variable drawn from a beta distribution. Allowing $p_i$ to be a random variable helps statistically account for the the confounding effects of variable diversification rate and lineage age on lineage richness [@mcpeek2007; @scholl2016; @yu2024]. Drawing $p_i$ from a beta distribution averages over these possible confounding effects.

The resulting beta-geometric distribution has parameters $\alpha$ and $\beta$: 

\begin{equation}
P(y; \alpha, \beta) = \frac{B(\alpha + 1, y + \beta - 1)}{B(\alpha, \beta)}
\end{equation}

where $B$ is the beta function.

Interestingly enough, the Yule-Simons distribution, historically developed to model number of species in genera [@yule], is a special case of the beta-geometric distribution when $\beta = 1$ [@irwin1975]. In our beta-geometric model we will not fix $\beta = 1$ but instead allow it to be a free parameter, thus not making the strict birth-only assumption of @yule

Our final, quadratic model relating genus richness to average per-species abundance in a genus is


\begin{align}
y_i &\sim \text{BetaGeometric}(\alpha_i, \beta) \\
\log(\alpha_i) &= a_0 + a_1 x_i + a_2 x_i^2 \label{eq-link} \\
\log(\beta) &= b_0
\end{align}


where $x_i$ is average per-species abundance of lineage $i$. This model means that we allow the $\alpha$ parameter of the beta-geometric distribution to change with average per-species abundance but we assume the $\beta$ parameter is fixed across lineages. A fixed $\beta$ across lineages still allows sufficient flexibility of the model without introducing additional parameters for which there may be insufficient power to reliably estimate. From this model our estimate for the observable diversification probability for lineage $i$ is $1 - \hat{p_i} = 1 - \frac{\alpha_i}{\alpha_i + \beta}$.

To evaluate the significance of an intermediate peak in diversification across average per-species abundances (i.e. the quadratic term) we also fit a model without quadratic terms:


\begin{align}
y_i &\sim \text{BetaGeometric}(\alpha_i, \beta) \\
\log(\alpha_i) &= a_0 + a_1 x_i \\
\log(\beta) &= b_0
\end{align}


We then compare quadratic and linear models with a likelihood ratio test to evaluate significance of the intermediate peak.

While the *VGAM* package [@vgam] in R enables beta-geometric generalized linear modeling, its implementation is based on iterative Fisher scoring which is unnecessary and computationally slow for our application. Therefore, we use functionality from *VGAM* to write a custom likelihood maximization routine that we use to estimate model parameters. More details on this approach are provided in \ssec{sec-bgeom}.



## Results 

### Simulating a birth-death-immigration process with speciation

\ssec{sec-sim-res} provides details on the outcomes of the simulation experiment, in particular documenting the effect of differing values for speciation rate, dispersal rate, and waiting time to full speciation

### Simulation experiment: effect of abundance on speciation

```{r}
load_supp("sim-mod")
```

In visualizing simulation runs that did and did not produce speciation events, we can see a clear peak in the probability of speciation at intermediate abundance [@fig-sim]. A logistic regression model with a quadratic term confirms this peak, finding that the coefficient on the quadratic term falls within the 95% confidence interval (`r round(sim_quad_ci[1], 3)`, `r round(sim_quad_ci[2], 3)`)

```{r}
#| label: fig-sim
#| fig-width: 4.5
#| fig-height: 3
#| fig-cap: |
#|     Relationship across simulation run windows between average abundance 
#|     and speciation across. Histograms across bottom and top 
#|     show the frequencies of windows without speciation (bottom) and with 
#|     full-speciation (top) across different different incipient speciation 
#|     rates. Curve is a binomial generalized linear model and shaded region 
#|     shows 95% confidence envelope.

# loads `sim_fig`
load_supp("make-sim-fig")

sim_fig
```

### Relationship between genus abundance and species richness in Hawai\`i arthropods

```{r}
load_supp("arth-data")
load_supp("arth-mod")
load_supp("lrt")
load_supp("make-fig-arth")

arth_mod_ci <- confint(arth_mod)
```

Arthropods in Hawai\`i show a hump-shaped relationship between abundance and species richness [@fig-arth]. There is a clear outlier with higher genus species richness (above 300) than expected based on abundance. This outlier is *Drosophila*, which is not surprising given both the extensive systematics work on Hawaiian *Drosophila* and their unique speciation mechanisms [@ogrady2018], both of which could inflate the group's reported species richness. 

The hump-shaped relationship is captured by a quadratic generalized beta-geometric model with sample size $n = `r nrow(gen)`$ and coefficients *intercept =* `r format(round(arth_mod$coefficients[1], 3), nsmall = 3)` (95% CI: [`r paste(format(round(arth_mod_ci[1, ], 3), nsmall = 3), collapse = ", ")`]), *linear term =* `r format(round(arth_mod$coefficients[2], 3), nsmall = 3)` (95% CI: [`r paste(format(round(arth_mod_ci[2, ], 3), nsmall = 3), collapse = ", ")`]), *quadratic term =* `r format(round(arth_mod$coefficients[3], 3), nsmall = 3)` (95% CI: [`r paste(format(round(arth_mod_ci[3, ], 3), nsmall = 3), collapse = ", ")`]). 95% CIs are  Wald confidence intervals. Coefficient values are reported on the link scale (i.e. as in eq. \ref{eq-link}).

A likelihood ratio test comparing quadratic and linear models of genus richness across average per-species abundance favors the quadratic model with a likelihood ratio of `r round(arth_lrt$Chisq[2], 3)` and a *P*-value of `r round(arth_lrt[["Pr(>Chisq)"]][2], 5)` from a $\chi^2$ distribution with $d.f.$ = `r arth_lrt$Df[2]`. Removing *Drosophila* has no meaningful impact on the relationship or the statistical significance of the quadratic term (see \ssec{sec-bgeom}).


```{r}
#| label: fig-arth
#| fig-width: 7
#| fig-height: 3
#| fig-cap: |
#|     Relationship between genus abundance and species richness for arthropods 
#|     endemic to Hawai\`i. Panel A shows the fitted relationship between 
#|     abundance and the observable diversification probability $1 - \hat{p}$.
#|     Solid line is expected value and gray shaded region is 95% confidence 
#|     envelope. Panel B shows the pattern in the raw data along with the 
#|     relationship predicted by the quadratic model. The solid black line
#|     represents the expected value and gray shaded regions the 95% confidence
#|     envelope. Colors of points reflect the local density of data points such 
#|     that a yellow point indicates a cluster of greater than 10 data points at
#|     that specific x-y location and dark a purple point indicates only a 
#|     single data point. The outlier with genus richness above 300 is 
#|     *Drosophila* discussed further in the main text and \ssec{sec-bgeom}

gp_arth
```

## Discussion

Taken all together these results indicate support, from both theoretical and empirical perspectives, for the idea that intermediate abundances promote speciation. Using a theoretical simulation-based model we show that the chance of speciation peaks at intermediate abundances because these intermediate abundances represent a balance between dispersal, which slows speciation, and population persistence plus per capita incipient speciation, which are necessary for speciation. The balance arises because dispersal is lowest at low abundance (meaning the rate of transition from incipient to full speciation is greatest at low abundance) while persistence and incipient speciation are highest at high abundance. Our work suggests an intriguing possibility that if properties like abundance and dispersal are emergent species-level traits that modulate speciation, they could be acted upon by macroevolutionary selection [@rabosky2010]
.
Our results are novel in that they go against the widespread assumption that more abundant lineages should be those that produce more new speciation---an assumption tracing back to @darwin and also integral to the unified neutral theory of biodiversity [@hubbell2001]. However, our theoretical results are also grounded in published hypotheses, specifically the role of dispersal in modulating speciation [@price2004; @claramunt2012; @czekanski2019; @yamaguchi2022].

Our empirical results are qualitatively consistent with the theoretical finding that speciation peaks at intermediate abundances. The empirical results are made possible by the unique evolutionary setting of the Hawaiian Archipelago, where extreme isolation aids in the direct observation of diversification [@wagner-funk1995; @price2004]. However, inferring past diversification from modern-only data has been shown to have multiple technical pitfalls including artifacts of sampling [@gruner2008], model non-identifiability [@louca2020], and difficulty in dissambiguating effects of clade age and clade diversification rate [@Wiens]. Thus, we emphasize that the empirical results should be interpreted in concert with the theoretical results, each adding distinct and complementary support to the idea of peak speciation at intermediate abundances.

The empirical and theoretical results are also distinct and cannot be interpreted as quantitatively equivalent. While the x-axis locations of the quadratic curve peaks in @fig-sim and @fig-arth are roughly similar, this does not mean that the simulation model parameters have a one-to-one mapping onto the parameters of a hypothetical diversification model that could have produced the empirical patterns. The parameters of the simulation model have to be interpreted separately from the empirical results because of key differences between the simulation set-up and the empirical analysis. Firstly, the "response variables" of each analysis are different---the simulation is able to directly observe speciation and thus we directly analyze speciation probability; conversely, we cannot directly observe speciation in the empirical system and so instead model $1 - \hat{p}$, the "observable diversification probability." The fact that we stop a simulation run once speciation is achieved also makes the simulated results different from the empirical system---in the real world, time does not stop if speciation occurs; instead, some clades continue to diversify and some go extinct. Our simulation model is thus one of the probability of first speciation and does not account for how the probability of speciation may itself change over the course of diversification [e.g., @rabosky2009; @wagner2014; @calcagno2017; @foote2023]. The continued diversification of real world clades could also explain the flatter relationship of $1 - \hat{p}$ across abundances. Additionally, while both the simulation model and empirical analysis are concerned with abundance, the way abundance is recorded in each is different. In the simulation we know the complete population size of every local population and we scale rates such that local population sizes are computationally tractable. Real world population sizes (perhaps up to $10^8$) would not be computationally manageable. Also unlike the simulations, the real world data represent tiny subsamples of the whole populations.

Nonetheless, the simulation model provides a proof-of-concept and a mechanism for how speciation might depend on abundance in a more nuanced way than previously considered. Critically, our model does not depend on a neutrality assumption to arrive at this conclusion. Although our model is inspired by the agent-based modeling approaches of neutral theories [@hubbell2001; @rosindell2013; @rosindell2015] we only model populations of a single species and thus make no assumption about how multiple species interact or whether different species have different eco-evolutionary rates. Thus our findings apply regardless of whether one assumes neutral- or niche-based dynamics. We also do not assume density dependence, as this is not our focus. However, can postulate that positive density-dependent dispersal [@travis1999], which causes total immigration rate to increase super-linearly with abundance, could further accentuate our findings. Conversely, negative density-dependent dispersal [@rodrigues2014] could dampen the effects we find, but, so long as per-capita dispersal remains a non-zero probability, likely not eliminate them.


Because our model focuses on a single species and thus makes no assumptions about dynamics or interactions between species we highlight the unnecessary assumptions of previous approaches. @darwin, for example, was also interested in the relationship between abundance and speciation. He assumed that more abundant species gave rise to more new species because those abundant species were biologically dominant to less abundant one, and that this dominance promoted both abundance and diversification [@darwin; @maurer1999]. The language of @darwin, including reference to "conquer[ing]" [p. 125] and "triumphant" [p. 126] clades, cannot be separated from discourse then and now that seeks to justify racism and colonial violence [@brantlinger2003; @hesse2007; @fuentes2021; @diogo2023]. Beyond the inherent amorality of such assumptions, our work shows they are unnecessary in explaining why abundance connects to speciation. We also show that these unnecessary assumptions lead to a prediction---a positive relationship between abundance and speciation---that may not universally hold.






## References 
